---
title: "Using the t-test"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NULL)
```

## 7.1 One-sample  means with the *t*-distribtion


```{r}
x <- seq(-4, 4, length.out = 200)
dnormx <- dnorm(x)
dtx <- dt(x, 4)
plot(dnormx ~ x, type = "l", ylab = "")
lines(x, dtx, lty = 2)
legend("topright", c("Normal", "t"),lty = c(1,2))
```

#### Example 7.2

What proportion of the t-distribution with 18 degrees of freedom falls below -2.10?

```{r}
pt(q = -2.10, df = 18)
```


#### Example 7.3

A t-distribution with 20 degrees of freedom is shown in the left panel of Figure 7.4. Estimate the proportion of the distribution falling above 1.65.

```{r}
pt(q = 1.65, df = 20, lower.tail = FALSE)
```

#### Example 7.4

A t-distribution with 2 degrees of freedom is shown in the right panel of Figure 7.4. Estimate the proportion of the distribution falling more than 3 units from the mean (above or below).

```{r}
pt(3, 2, lower.tail = FALSE) +  pt(-3, 2)
# or
pt(3, 2, lower.tail = FALSE)*2
```

EXAMPLE 7.6
Are the independence and normality conditions satisfied for this data set?



The observations are a simple random sample, therefore independence is reasonable. The summary
statistics in Figure 7.6 do not suggest any clear outliers, since all observations are within 2.5 standard
deviations of the mean. Based on this evidence, the normality condition seems reasonable.
In the normal model, we used z? and the standard error to determine the width of a confidence
interval. We revise the confidence interval formula slightly when using the t-distribution:


point estimate $\pm t^*_{df} \times  SE$ →  $\bar{x} \pm t^*_{df} \times \frac{s}{\sqrt{n}}$

EXAMPLE 7.7
Using the summary statistics in Figure 7.6, compute the standard error for the average mercury
content in the n = 19 dolphins.

We plug in s and n into the formula: $SE = \frac{s}{\sqrt{n}} = \frac{2.3}{\sqrt{19}} = 0.528$.

```{r}
xbar <- 4.4
s <- 2.3
n <- 19
SE <- s/sqrt(n)
SE
```

## EXAMPLE 7.8

When n = 19, what is the appropriate degrees of freedom? Find $t^*_{df}$ for this degrees of freedom and
the confidence level of 95%

The degrees of freedom is easy to calculate: df = n − 1 = 18.
Using statistical software, we find the cutoff where the upper tail is equal to 2.5%: $t^*_{18} = 2.10$. The
18area below -2.10 will also be equal to 2.5%. That is, 95% of the t-distribution with df = 18 lies
within 2.10 units of 0.

## EXAMPLE 7.9

Compute and interpret the 95% confidence interval for the average mercury content in Risso’s
dolphins.
We can construct the confidence interval as
x̄ ± t? 18 × SE
 →
 4.4 ± 2.10 × 0.528
 →
 (3.29, 5.51)

```{r}
c <- qt(.975, df = n-1)
CI.lwr <- xbar - c*SE
CI.upr <- xbar + c*SE
cat("Confidence interval for x = (", round(CI.lwr, 2), ", ", round(CI.upr,2), ")", sep = "")
```


 