---
title: "Using `ggplot2`"
author: "William Murrah"
date: "5/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Return to Tutorial Menu](https://auqmer.github.io/TutoR/)

## Getting Started

The `ggplot2` package is a very powerful graphics package created by Hadley Wickham. 
It is not part of base R, so you will need to install it if you have not already done so.

```{r, eval=FALSE}
install.packages("ggplot2")
```

Once it is installed, you can load the package from your R library as follows:
```{r}
library(ggplot2)
```

## Overview

`ggplot2` is a graphics package based on *The Grammer of Graphics*, a book by Leland Wilkinson that presents a unique way of producing graphics. 
Creating plots in `ggplot2` is different the using base graphics, but once you get used to it you will most likely find it very useful. 
This tutorial is just a basic introduction to the package.
To find out more visit the package's webpage at https://ggplot2.tidyverse.org/.

For this tutorial we will use the `mtcars` data, which is part of the base R datasets.
You can load it with the `data()` function which I do below. 
I also convert some of the variables to factors. 
You should also open the help page for the dataset by typing `?mtcars` into the R console.


```{r}
data(mtcars)

mtcars <- within(mtcars, {
  cyl <- factor(cyl)
  vs <- factor(vs, labels = c("V-shaped", "straight"))
  am <- factor(am, labels = c("automatic", "manual"))
})
```

```{r}
ggplot(mtcars, aes(hp, mpg, color = cyl)) + geom_point()
```



[Return to Tutorial Menu](https://auqmer.github.io/TutoR/)